---
title: "Reproducible Research Week 2"
author: "JAGV"
date: "2023-04-19"
output: html_document
---

## Setting global options

```{r setoptions, echo = FALSE}
## knitr::opts_chunk$set(echo = TRUE, results = "hide")
```


# Coding Standars for R

1. Always use text files/text editor.
2. Indent your code.
    + Indenting improves readability.
    + Fixing line length (80 columns) prevents lots of nesting and very long functions.
    + Suggested: Indents of 4 spaces at minimum; 8 spaces ideal.
3. Limit the width of your code (80 columns?)
4. Limit the length of individual functions.

# Markdown

Markdown is a text-to-HTML conversion tool for web writers. Markdown allows you to write using an easy-to-read, easy-to-write plain text format, then convert it to structurally valid XHTML (or HTML).

Some examples include:

- Italics: *This text will appear italicized!*
- Bolds: **This one will appear in bold font**
- Headers: Each $#$ sign will indicate a header.
- Links: $[text](url)$ --> [RStudio](http://rstudio.com/)

#### Advanced linking

There is an example of advanced linking. The page with several R apps [The Art of Stat][1] or the page with [a data set on Pokémons][2]

[1]: https://artofstat.com/web-apps
[2]: https://www.kaggle.com/datasets/abcsds/pokemon

# R Markdown

R Markdown can be converted to standard markdown using the `knitr` package in R, and can be converted to HTML using the `markdown` package in R.


Bonus: slides can be created in R Markdown and converted to slides using the `slidify` package!

# R Markdown demonstration


```{r}
library(datasets)
data(airquality)
summary(airquality)
```

Let's make a pairs plot of the data.

```{r}
pairs(airquality)
```
Here's a regression model of Ozone on Winf, Solar radiation, and temperature.

```{r}
library(stats)
fit <- lm(Ozone ~ Wind + Solar.R + Temp, data = airquality)
summary(fit)
```

# Literate Statistical Programming with knitr

How do I make my work reproducible?

- Decide to do it (ideally from the start)
- Keep track of things, perhaps with a version control system to track snapshots/changes
- Use software whose operation can be coded.
- Do not save output.
- Save data in non-proprietary formats.

Some pros and cons of literate programming.

Pros:

- Text and code all in one place, logical order.
- Data, results automatically updated to reflect external changes.
- Code is live --  automatic "regression test" when building a document.


Cons:

- Text and code all in one place, can make documents difficult to read, especially if there is a **lot** of code.
- Can substantially slow down processing of documents (although there are tools to help).

## In-line text computations

```{r date, echo = FALSE}
time <- format(Sys.time(), "%a %b %d %X %Y")
rand <- rnorm(1)
```

The current time is `r time` and a normal standard random number is `r rand`.

## Incorportating graphics

Let's first simulate some data.
```{r simdata, echo = TRUE}
x <- rnorm(100)
y <- x + rnorm(100,sd=0.5)
```

Here is a scatterplot of the data.

```{r scatterplot, fig.height = 4}
par(mar=c(5,4,1,1),las = 1)
plot(x,y,main = "My Simulated Data")
```

## Making tables with xtable

```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone ~ Wind + Temp + Solar.R, data = airquality)
```

Now, here you have a table with the regression coefficients

```{r showtable, results = "asis"}
if(!require("xtable")) install.packages("xtable")
library(xtable)
xt <- xtable(summary(fit))
print(xt,type = "html")
```

## Caching computations

One chunk may take a long time to run, and as all chunks have to be re-computed every time we re-knit the file, then the `cache = TRUE` option can be set in order to store results of computation. After the first run, results are loaded from cache.

Warning! Chunks with significant *side effects* may not be cacheable.

## Summarizing

To sum all up, **literate statistical programming** can be a useful way to put, text, code, data, output all in one document, and `knitr` is a powerful tool for integrating code and text in a simple document format.





