---
title: "R Notebook"
author: "Jorge Garza-Venegas"
date: "2022-07-25"
output: html_notebook
---

# Week 3: Loop Functions and Debugging

## Loop functions - lapply and sapply

There are several loop functions in R, often with the word "apply" in their names. For instance, 

- `lapply`: loops over a **list** and evaluate a function on each element.  
- `sapply`: same as `lapply` but try to simplify the result.  
- `apply`: Apply a function over the margins of an array.  
- `tapply`: Apply a function over subsets of a vector.  
- `mapply`: Multivariate version of `lapply`.

An auxiliary function `split` is also useful, particularly in conjunction with `lapply`.

```{r}
lapply
```

**lapply** always returns a list.

```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x,mean)

x <- list(a = 1:4, b = rnorm(10), c = rnorm(20,1), d = rnorm(100,5))
lapply(x,mean)
```

More examples

```{r}
x<- 1:4
lapply(x,runif)

x <- 1:4
lapply(x, runif, min =0, max = 10)
```

You can also use *anonymus functions* within `lapply`. The follwing example will extract the first column of two matrices in a list.

```{r}
x <- list(a = matrix(1:4, 2,2), b = matrix(1:6, 3, 2))
x
lapply(x, function(elt) elt[,1])
```

**sapply** try to simplify the results in the following way:

- if the result is a list which every element is of length 1, then a vector is returned.  
- if the results is a list which every element is of length $>1$, then a matrix is returned.  
- if it can not figure it out, a list is returned.

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20,1), d = rnorm(100,5))
lapply(x,mean)
sapply(x,mean)

## It is not smart to use mean on x, since x is a list and not a vector.
mean(x)
```

## Loop functions - apply

`apply` is used to evaluate a function over the margins of an array:

- Rows and columns of a matrix.  
- General arrays.  

Despite it is not really faster than writing a loop, it can be written in just one line.

```{r}
str(apply)
```

The parameters of `apply` are: 

- X, an array.  
- MARGIN, an integer vector indicating the margins that will be retained.  
- FUN, the function to be applied.  
- ..., more arguments.

```{r}
x <- matrix(rnorm(200), 20, 10)
## Calculate the mean column-wise (the second dimension).
apply(x,2,mean)
## Calculate the sum of the rows (the first dimension).
apply(x,1,sum)
```

Shortcuts for some common used functions over matrices:

- `rowSums` equals to `apply(x,1,sum)`  
- `rowMeans` equals to `apply(x,1,mean)`  
- `colSums` equals to `apply(x,2,sum)`  
- `colMeans` equals to `apply(x,2,mean)`  

they are *much* faster than the `apply`, which will be noticeable for large matrices.

Other types of functions that can be used with **apply**

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(0.25, 0.75))
## The quantile functions needs the probs argument, so it is mandatory to pass this argument via the apply "..." argument.
```

Note that the quantiles are returned in a matrix, with 2 rows and 20 columns.


### Average matrix in an array

```{r}
a <- array(rnorm(2*2*10), c(2,2,10))
apply(a, c(1,2), mean)
rowMeans(a,dims = 2)
```

## Loop Functions - mapply

In contrast with `lapply` and `sapply`, `mapply` can take several arguments (for example, more than just one list).

```{r}
str(mapply)
```

```{r}
list(rep(1,4),rep(2,3),rep(3,2),rep(4,1))
```

This can be done in a more compact way using `mapply`

```{r}
mapply(rep, 1:4,4:1)
```

### Vectorizing a function

```{r}
noise <- function(n,mean,sd){
    rnorm(n,mean,sd)
}
noise(5,1,2)
## This sentence is intended to create 1 random number with mean 1, 2 random numbers with mean 2, and so on. 
noise(1:5,1:5,2)
## But we do not have that result. Instead, we should use mapply
mapply(noise, 1:5,1:5,2)
```


## Loop functions - tapply

Loop over a subset of a vector. This can be helpful for data summary within factors (gender, preferences, etc).

```{r}
str(tapply)
```

Let's take a look at an example, where we will use the function `gl` to create 3 factors (levels) of the same size.

```{r}
x <- c(rnorm(10), runif(10), rnorm(10,1))
## Create factors.
f<-gl(3,10)
f
## Calculate the mean of X within the f factors (levels)
tapply(x,f,mean)
tapply(x,f,mean, simplify = FALSE)
```

For the range functions within groups

```{r}
tapply(x,f,range)
```

## Loop functions - split

`split` takes an R object (usually a vector) and splits it into groups determined by a factor or list of factors. It is not a loop function, but it is often used with loop funcionts such as `lapply` and `sapply`. It has the following arguments

```{r}
str(split)
```

where the argument **drop** is used to indicate whether or not empty factors should be dropped.

```{r}
x <- c(rnorm(10), runif(10), rnorm(10,1))
## Create factors.
f<-gl(3,10)
## Split the vector into 3 vectors: one for each factor.
split(x,f)
```

One example of the conjoint use of `split` and `lapply`

```{r}
lapply(split(x,f),mean)

## Which is the same as using the tapply function
tapply(x,f,mean)
```

Another example with a more complicated dataset.

```{r}
library(datasets)
head(airquality)
```

We might want to calculate summary statistics within months. We can do that using `split`.

```{r}
s<-split(airquality, airquality$Month)
lapply(s,function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")]))
sapply(s,function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")]))
sapply(s,function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")], na.rm = TRUE))
```

Splitting in more than one level. We can use the function `interaction` to combine all levels of each factors to create a new vector of factors.

```{r}
x <-rnorm(10)
x
f1 <- gl(2,5)
f2 <- gl(5,2)
f1
f2
interaction(f1,f2)
```

Interactions can create empty levels.

```{r}
str(split(x,f1))
str(split(x,f2))
str(split(x,list(f1,f2)))
str(split(x,list(f1,f2), drop = TRUE))
```





