---
title: "Simulation and profiling"
author: "Jorge A. Garza-Venegas"
date: "29/7/2022"
output: html_document
---

## The `str` function

This function is used to display compactly the internal **structure** of an R object.

```{r}
str(str)

str(lm)

x <- rnorm(100,0,2)
summary(x)
str(x)

f <- gl(40,10)
str(f)
```

We can use this function with data frames to check the types of their column variables.

```{r}
head(airquality)
str(airquality)
```

```{r}
s <- split(airquality, airquality$Month)
summary(s)
str(s)
```

## Simulation - Generating random numbers.

There are several functions in R to generate random variables, following the structure

\[
\textit{prefix} + \textit{suffix}  
\]
where **prefix** can be one of the following letters:

- **d** for density.  
- **r** for random number generation.  
- **p** for cumulative distribution.  
- **q** for quantile function.  

and **suffix** is a distribution name. Some examples are 

+----------+-----------------------------+
| suffix   | Distribution Name           |
+----------+-----------------------------+
| binom    | Binomial distribution       |
+----------+-----------------------------+
| geom     | Geometric distribution      |
+----------+-----------------------------+
| pois     | Poisson distribution        |
+----------+-----------------------------+
| norm     | Normal distribution         |
+----------+-----------------------------+
| exp      | Exponential distribution    |
+----------+-----------------------------+
| gamma    | Gamma distribution          |
+----------+-----------------------------+
| weibull  | Weibull distribution        |
+----------+-----------------------------+
| t        | t Distribution              |
+----------+-----------------------------+
| f        | F distribution              |
+----------+-----------------------------+
| chisq    | chi-squared distribution    |
+----------+-----------------------------+

Let's take a look at some examples

```{r}
# Simulating random normal variables.
x <- rnorm(10)
x


y<- rnorm(n = 10, mean = 20, sd = 2)
y
summary(y)
```

Setting a random number seed **ensures reproducibility**. In R, this can be done with the function `set.seed()`

```{r}
rnorm(5)
rnorm(5)
set.seed(1)
rnorm(5)
set.seed(1)
rnorm(5)
rnorm(5)
```

```{r}
# Exponential distribution
rexp(n = 10, rate = 1)
# P(X <=1)
pexp(q = 1, rate = 1)

# Binomial distribution
rbinom(n = 2, size = 10, prob = 0.5)
# P(X < 5)
qbinom(0.5, size = 10, prob = 0.5)
qbinom(0.6230469, size = 10, prob = 0.5)
pbinom(6,10,0.5) - pbinom(5,10,0.5)
dbinom(6,10,0.5)
```

## Simulation - Simulating a linear model

In order to model the linear model given by 

\[
y = \beta_0 + \beta_1x + \varepsilon
\]
where $\varepsilon \sim \mathcal{N}\left(0,\sigma^2\right)$ and assuming $\beta_0 = 0.5$ and $\beta_1 = 5$.

```{r}
beta_0 <- 0.5
beta_1 <- 5
set.seed(20)
x <- rnorm(100)
e <- rnorm(100,0,2)
y <- beta_0 + beta_1*x + e
summary(y)
plot(x,y)
```

Generate binary data using a Binomial Distribution.

```{r}
beta_0 <- 0.5
beta_1 <- 5
set.seed(20)
x <- rbinom(100, 1,0.5)
e <- rnorm(100,0,2)
y <- beta_0 + beta_1*x + e
summary(y)
plot(x,y)
```

Another example. Assume that $Y \sim Poisson\left(\mu\right)$ and that $\log\left(\mu\right) = \beta_0 + \beta_1 x$ with $\beta_0 = 0.5$ and $\beta_1 = 0.3$.

```{r}
set.seed(1)
beta_0 <- 0.5
beta_1 <- 0.3
x <-rnorm(100)
log.mu <- beta_0 + beta_1*x
y <- rpois(n = 100, exp(log.mu))
summary(y)
plot(x,y)
```

## Simulation - Random Sampling

One useful function is the `sample` function.

```{r}
set.seed(1)
sample(1:10,4)
sample(1:10,4)
sample(letters,5)
sample(1:10)  # This is a permutation.

sample(1:10, replace = TRUE)
```

## R Profiler (Part 1)

**Profiling** is a systematic way to examine how much time is spend in different parts of a program and it is very useful for code optimization.

General principles of optimization

- Design, first, then optimize.  
- "Premature optimization is the root of all evil" Donald Knuth.  
- Measure (collect data), do not guess.  

### The `system.time()` function

This function computes the time (in seconds) needed to execute an expression as it can be seen in the following examples.

```{r}
system.time(rnorm(100))

system.time(rnorm(100000000))
```

where **user** is the time charged to the CPU(s) for this expression and **elapsed** is the time experienced by the user and sometimes called "wall clock" time. **Elapsed** might be smaller than **user** when your computer has multiple processors and is capable of use them. (One way to use them is via the **parallel** package)

```{r}
system.time(readLines("http://www.jhsph.edu"))

hilbert <- function(n){
    i <- 1:n
    1/outer(i-1,i,"+")
}
x <- hilbert(1000)
system.time(svd(x))
```

Longer expression can be timed using curly braces.
```{r}
system.time({
    n <- 100
    r <- numeric(n)
    for (i in 1:n) {
        x <-rnorm(n)
        r[i] <- mean(x)
    }
})
```

This is useful when you now where the problem is, or at least have an idea where to check.

## R Profiler (Part 2)

The profiler in R is started with the `RProf()` function. It keeps track of the function call stack at regularly sampled intervals (default is 0.02 secs) and tabulates how much time is spend in each function.

The output of this function is usually "not readable", so that the `summaryRProf()` might be used. 

You have to note that C or Fortran code are not profiled.








