---
title: "C8 Week 4"
author: "JAGV"
date: "2026-06-23"
output: html_document
---
# Regularized regression

Basic idea: fit a regression model and penaliza (or shrink) large coefficients. 

**Pros**

- Can help with the bias/variance tradeoff
- Can help with model selection


**Cons**

- May be computationally demanding on large data sets
- (In general) Does not perform as well as random forests and boosting.



### A motivating example

Assume the regression model

\[
    Y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \varepsilon
\]

where $x_1$ and $x_2$ are nearly perfectly correlated (co-linear), Then, you can approximate this model by:

\[
    Y = \beta_0 + \left(\beta_1 + \beta_2\right)x_1 + \varepsilon
\]


The result is: 

- You will get a good estimate of $Y$.
- The estimate of $Y$ will be biased.
- We may reduce variance in the estimate.


Let's take a look at one example: `prostate cancer`

```{r}
prostate <- read.csv("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data", sep="\t",head=T,row.names=1)
str(prostate)
```
**Note** the code for this example comes from *Hector Corrada Bravo's Practical Machine Learning Course* to demonstrate the behaviour of training and test errors, the `prostate` dataset from *Elements of Statistical Learning* is used.

```{r}
set.seed(1, sample.kind = "Rounding")
# Define outcome "y" and predictors "x".
covnames <- names(prostate[-(9:10)])
y <- prostate$lpsa
x <- prostate[,covnames]
# Create test set predictors and outcomes.
train.ind <- sample(nrow(prostate), ceiling(nrow(prostate))/2)
y.test <- prostate$lpsa[-train.ind]
x.test <- x[-train.ind,]
# Create training set predictors and outcomes
y <- prostate$lpsa[train.ind]
x <- x[train.ind,]
# Define p, the number of predictors
p <- length(covnames)
# Initialize the list of residual sum squares
rss <- list()
# Loop through each combination of predictors and build models
for (i in 1:p) {
    # Compute matrix for p choose i predictors for i = 1...p (i \times p matrix)
    Index <- combn(p,i)
    # Calculate RSS of each combination of predictors.
    rss[[i]] <- apply(Index, 2, function(is) {
        # Take each combination (or column of Index matrix) and create formula for regression
        form <- as.formula(paste("y~", paste(covnames[is], collapse="+"), sep=""))
        # run LR with combination of predictors on training data
        isfit <- lm(form, data = x)
        # predict outcome for all training data points
        yhat <- predict(isfit)
        # Calculate RSS for predictions on Training data.
        train.rss <- sum((y-yhat)^2)
        # Predict outcome for all test data points
        yhat <- predict(isfit, newdata = x.test)
        # Calculate RSS for predictions on test data
        test.rss <- sum((y.test - yhat)^2)
        # Store each pair of training and test residual sum of squares as a list
        c(train.rss, test.rss)
    })
}

# Set up plot
plot(1:p, 1:p, type = "n", ylim = range(unlist(rss)), xlim = c(0,p), 
     xlab = "Number of predictors", ylab = "RSS", 
     main = "Prostate Cancer Data - Training vs Test RSS")
# Add data points for training and test RSS
for( i in 1:p) {
    # Plot training residual sum squares in blue
    points(rep(i, ncol(rss[[i]])), rss[[i]][1,], col = "blue", cex = 0.5)
    # Plot test residual sum squares in red
    points(rep(i, ncol(rss[[i]])), rss[[i]][2,], col = "red", cex = 0.5)
}
# Find the minimum training RSS for each combination of predictors
minrss <- sapply(rss, function(x) min(x[1,]))
#Plot line through the minimum training RSS data points in blue
lines((1:p), minrss, col = "blue", lwd = 1.7)
# Find the moinimum test RSS for each combination of predictors 
minrss <- sapply(rss, function(x) min(x[2,]))
#Plot line through the minimum test RSS data points in red
lines((1:p), minrss, col = "red", lwd = 1.7)
# Add legend
legend("topright", c("Train", "Test"), col = c("blue", "red"), pch = 1)
```

From here, we can see that the train set error always goes down, whereas the test set error exhibits a typical pattern: it goes down, hit a plateau and then goes back up again. This is because we are overfitting the data in the training set. 


## Model selection approach: 

### Split samples

No method better when data/computation time permits it. The approach is as follows:

- Divide data into training/test/validation sets
- Treat validation as test data, train all competing models on the trin data and pick the best one on validation
- To appropriately assess performance on new data apply to test set
- You may re-split and reperform steps 1-3

Two common problems

- Limited data
- Computational complexity


### Decomposing expected prediction error

Assume $Y_i = f\left(X_i\right) + \varepsilon_i$, then the expected prediction error is given by $EPE\left(\lambda\right) = E\left[\left\{Y - \hat{f}_{\lambda}\left(X\right)\right\}^2\right]$. 

Suppose that $\hat{f}_{\lambda}$ is the estimate from teh training data and look at a new data point $X = x^*$. Then 

\[
    E\left[\left\{Y - \hat{f}_{\lambda}\left(x^+\right)\right\}^2\right] = \underbrace{\sigma^2}_{\text{Irreducible}} + \underbrace{\left\{E\left[\hat{f}_{\lambda}\left(x^+\right)\right] - f\left(x^+\right)\right\}^2}_{\text{Bias}^2} + \underbrace{var\left[\hat{f}_{\lambda}\left(x_0\right)\right]}_{\text{Variance}}
\]

### Hard thresholding

When having less observations (samples) than covariates, the design matrix will be singular. Therefore, as a solution to this, for the model $Y = f\left(X\right) + \varepsilon$ set $\hat{f}_{\lambda}\left(x\right) = x'\beta$ and constrain only $\lambda$ of the coefficients to be nonzero. 

Selection problem is after choosing $\lambda$ figure out which $p-\lambda$ coefficients to make nonzero.

#### Regularization for regression


If the $\beta_j$'s are unconstrained:

- They can explode
- And hence are susceptible to very high variance 

To control variance, we might regularize/shrink the coefficients

\[
    PRSS\left(\beta\right) = \sum_{j=1}^{n}{\left(Y - \sum_{i=1}^{m}{\beta_{1i}X_{ij}}\right)^2} + P\left(\lambda;\beta\right)
\]
where $PRSS$ is a penalized form of the sum of squares. Things that are commonly looked for:

- Penalty reduces complexity and variance
- Penalty respects structure of the problem


### Ridge regression

The problem is stated as solving

\[
    \sum_{i=1}^{N}{\left(y_i - \beta_0 - \sum_{j=1}^{p}{x_{ij}\beta_j}\right)}^2 + \lambda\sum_{j=1}^{p}\beta_j^2
\]
which is equivalent to solving

\[
    \sum_{i=1}^{N}{\left(y_i - \beta_0 - \sum_{j=1}^{p}{x_{ij}\beta_j}\right)}^2
\]

subject to $\sum_{j=1}^{p}{\beta_j^2} \leq s$ where $s$ is inversely proportional to $\lambda$. Inclusion of $\lambda$ makes the problem non-singular even if $X'X$ is not invertible.


#### Tuning parameter $\lambda$

We can see that $\lambda$ controls the size of the coefficients and the amount of **regularization**. Moreover, as $\lambda\rightarrow 0$, we obtain the LS solution, whereas as $\lambda\rightarrow \infty$ we have $\hat\beta = 0$.

So, taking the parameter can be done with cross-validation or techniques that help to find the optimal tradeoff of bias for variance.



### Regularized Regression - LASSO (Least Absolute Shrinkage and Selection Operator) regression

It is similar to **ridge regression**, with slightly different penalty term:

\[
    \sum_{i=1}^{N}{\left(y_i - \beta_0 - \sum_{j=1}^{p}{x_{ij}\beta_j}\right)}^2 + \lambda\sum_{j=1}^{p}\left|\beta_j\right|
\]
which is equivalent to solving

\[
    \sum_{i=1}^{N}{\left(y_i - \beta_0 - \sum_{j=1}^{p}{x_{ij}\beta_j}\right)}^2
\]

subject to constraint $\sum_{j=1}^{p}{\left|\beta_j\right|} \leq s$ where $s$ is inversely proportional to $\lambda$. 
**NOTE**: LASSO effectively performs model selection (choose subset of predictors) while shrinking other coefficients, whereas **ridge regression** only shrinks the coefficients.



# Combining predictors

Some key ideas on this topic:

- You can combine classifiers by averaging/voting
- Combining classifiers improves accuracy
- Combining classifiers reduces interpretability
- Boosting, bagging, and random forests are variants on this theme


## Real-life examples.

### Netflix competition

[Netflix Prize](https://web.archive.org/web/20120507052942/http://www.netflixprize.com/leaderboard)

### Heritage health prize - Progress Prize 1

(broken links)

[Market Makers](https://kaggle2.blob.core.windows.net/wiki-files/327/e4cd1d25-eca9-49ca-9593-b254a773fe03/Market%20Makers%20-%20Milestone%201%20Description%20V2%201.pdf)

[Mestrom](https://kaggle2.blob.core.windows.net/wiki-files/327/09ccf652-8c1c-4a3d-b979-ce2369c985e4/Willem%20Mestrom%20-%20Milestone%201%20Description%20V2%202.pdf)


## The basic intuition behind majority vote

Suppose we have 5 completely independent classifiers. If accuracy of each is $70\%$, then, the **majority vote accuracy** (the probability of the majority of the models achieving $70\%$ at the same time) is:

\[
    \text{Majority vote accuracy, MVA} = \mathbb{P}\left(\text{3 correct, 2 wrong}\right) + \mathbb{P}\left(\text{4 correct, 1 wrong}\right) + \mathbb{P}\left(\text{5 correct}\right)
\]

therefore, under the independent assumption, this might be calculated using the binomial distribution. So


\[
    \text{MVA} = 10\left(0.7\right)^3\left(0.3\right)^2 + 5\left(0.7\right)^4\left(0.2\right) + \left(0.7\right)^5 \approx 83.7
\]

**Note**: using 101 classifiers, this quantity becomes $99.9\%$.

The approaches for combining classifiers includes:

- For **similar** classifiers: bagging, boosting and random forests
- For **different** classifiers: model stacking and model ensembling


Let's take a look at one example: Wage data

```{r}
library(ISLR)
data(Wage)
library(ggplot2)
library(caret)
Wage <- subset(Wage, select = -c(logwage))

# Create a building data set and validation set

inBuild <- createDataPartition(y = Wage$wage, p = 0.7, list = FALSE)

validation <- Wage[-inBuild, ]
buildData <- Wage[inBuild,]

inTrain <- createDataPartition(y = buildData$wage, p = 0.7, list = FALSE)
training <- buildData[inTrain,]
testing <- buildData[-inTrain,]
```

```{r}
dim(training)
dim(testing)
dim(validation)
```
Build two different models

```{r}
mod1 <- train(wage ~., method = "glm", data = training)
mod2 <- train(wage ~., method = "rf", data = training, trControl = trainControl(method = "cv"), number = 3)
```


Prediction on the testing set

```{r}
pred1 <- predict(mod1, testing)
pred2 <- predict(mod2, testing)
qplot(pred1,pred2, colour = wage, data = testing)
```

Now, lest fit a model that combines predictors

```{r}
predF <- data.frame(pred1,pred2, wage = testing$wage)
combModFit <- train(wage ~., method = "gam", data = predF)
combPred <- predict(combModFit, predF)
```

Look at the testing errors

```{r}
sqrt(sum((pred1 - testing$wage)^2))
sqrt(sum((pred2 - testing$wage)^2))
sqrt(sum((combPred - testing$wage)^2))
```

Now let's try it out on the validation set

```{r}
pred1V <- predict(mod1, validation)
pred2V <- predict(mod2, validation)
predVDF <- data.frame(pred1 = pred1V, pred2 = pred2V)
combPredV <- predict(combModFit, predVDF)
```

Look at the errors on validation

```{r}
sqrt(sum((pred1V - validation$wage)^2))
sqrt(sum((pred2V - validation$wage)^2))
sqrt(sum((combPredV - validation$wage)^2))
```
It is lower with the combined models even on the validation set.


## Final comments

Recall: **scalability matters**: One important point to keep in mind when doing model ensembling is that this can lead to increases in computational complexity. So, it turns out that even though **Netflix has paid a million dollars** to the team that won the prize, **the solution (the millionaire dollar solution) was never actually implemented** because it was too computationaly intensive to apply to specific data sets.


# Forecasting

Let's talk about forecasting, which is the name used for *predictions* in time series data. Financial data is a kind of time series data, as we can see in the following link: [Google Finance](https://www.google.com/finance/beta)

The main difference is that **data is time-dependent**. Here, we are going to look for some components: 

- Trends - long term increase or decrease.
- Seasonal patterns - patterns related to time (week, month, year, etc.)
- Cycles - patterns that rise and fall periodically

Therefore, the approach of *subsampling* into train/test data sets is more complicated. Note: a similar problem arises when dealing with **spatial data** for the same reasons: dependency between nearby observations and location specific effects. 

Typically, the goal is to predict one or more observations into the future (which in regression would be **extrapolation**). All standard predictions can be used (with caution/sense!).

Be careful with spurious correlation, as this kxcd joke shows with geographical maps: [http://xkcd.com/1138/](http://xkcd.com/1138/) (red areas are the high density in population places).


Let's take a look at an example.

```{r}
#install.packages("quantmod")
library(quantmod)

from.dat <- as.Date("01/01/08", format = "%m/%d/%y")
to.dat <- as.Date("12/31/13", format = "%m/%d/%y")
getSymbols("GOOG", src = "yahoo", from = from.dat, to = to.dat)
head(GOOG)
```

```{r}
mGoog <- to.monthly(GOOG)
googOpen <- Op(mGoog)
ts1 <- ts(googOpen, frequency = 12)
plot(ts1, xlab = "Years+1", ylab = "GOOG")
```


One way to analyze time series data, is to decompose into:

- **Trend**: Consistently increasing pattern over time.
- **Seasonal**: when there is a pattern over a fixed period of time that recurs. 
- **Cyclic**: when data rises and falls over non fixed periods. 

We can use the function `decompose` to decompose (duh!) a time series into its parts:

```{r}
plot(decompose(ts1), xlab = "Years+1")
```

### Training and test sets

```{r}
ts1Train <- window(ts1, start = 1, end = 5)
ts1Test <- window(ts1, start = 5, end = 7)
ts1Train
ts1Test
```

## Forecasting with simple moving average.

\[
    \hat{Y}_t = \dfrac{1}{2k+1}\sum_{j = -k}^{k}{Y_{t+j}}
\]

This is a moving average of order $k$, centered.

```{r}
# install.packages("forecast")
library(forecast)
plot(ts1Train)
lines(ma(ts1Train, order = 3), col = "red")
```

## Forecasting with exponential smoothing

Simple exponential smoothing. 

\[
    \hat{Y}_{t+1} = \alpha Y_t + \left(1-\alpha) \hat{Y}_t 
\]


```{r}
ets1 <- ets(ts1Train, model = "MMM")
fcast <- forecast(ets1)
plot(fcast)
lines(ts1Test, col = "red")
```

```{r}
accuracy(fcast, ts1Test)
```

# Unsupervised prediction

Supervised prediction is regarded as prediction where you know what the labels are. Unsupervised prediction is when you do not know the labels for prediction. In those cases, to build a predictor you

- Create clusters
- Name clusters
- Build predictor for clusters

so that, in a new data set, you predict clusters

## Iris data set example

```{r}
data(iris)
library(ggplot2)
library(caret)
inTrain <- createDataPartition(y = iris$Species, p = 0.7, list = FALSE)

training <- iris[inTrain,]
testing <- iris[-inTrain,]
dim(training)
dim(testing)
```
## Cluster with k-means

```{r}
kMeans1 <- kmeans(subset(training, select = -c(Species)), centers = 3)
training$clusters <- as.factor(kMeans1$cluster)
qplot(Petal.Width, Petal.Length, colour = clusters, data = training)
```

In this case, the clusters are closely related to the Species, but this is in general not the case. We can compare those clusters against the real labels

```{r}
table(kMeans1$cluster, training$Species)
```
### Build a predictor

```{r}
modFit <- train(clusters ~., data = subset(training, select = -c(Species)), method = "rpart")
table(predict(modFit, training), training$Species)
```

### Apply on a test set

```{r}
table(predict(modFit, testing), testing$Species)
```



