---
title: "C8 Week 2"
author: "JAGV"
date: "2026-01-13"
output: html_document
---

# The Caret Package

```{r}
#install.packages("caret", dependencies = TRUE)
#library(caret)
```

## Machine learning algorithms in R

- Linear discriminant analysis
- Regression
- Naive Bayes
- Support Vector Machines
- Classification and regression trees
- Random forests
- Boosting
- among others

```{r}
library(caret)
library(kernlab)
data(spam)

inTrain <- createDataPartition(y = spam$type, p = 0.75, list = FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
dim(testing)
```
This is an example of how to split data. Now, we can fit a model

```{r, warning = FALSE}
set.seed(32343)
modelFit <- train(type ~., data = training, method = "glm")
modelFit
```
We can take a look at the final model:

```{r}
modelFit$finalModel
```

and also make predictions

```{r}
predictions <- predict(modelFit, newdata = testing)
predictions
```

```{r}
confusionMatrix(predictions, testing$type)
```

# Data slicing

We can create data partition using the function `createDataPartition` in order to split data into training and testing sets. This is helpful in cross validation for instance.

```{r}
library(caret)
library(kernlab)
data(spam)

inTrain <- createDataPartition(y = spam$type, p = 0.75, list = FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```

We can also create K-folds:

```{r}
set.seed(32323, sample.kind = "Rounding")
folds <- createFolds(y = spam$type, k = 10, list = TRUE, returnTrain = TRUE)
sapply(folds,length)
folds[[1]][1:10]
```


```{r}
set.seed(32323, sample.kind = "Rounding")
folds <- createFolds(y = spam$type, k = 10, list = TRUE, returnTrain = FALSE)
sapply(folds,length)
folds[[1]][1:10]
```
Resampling

```{r}
set.seed(32323, sample.kind = "Rounding")
folds <- createResample(y = spam$type, times = 10, list = TRUE)
sapply(folds,length)
folds[[1]][1:10]
```
Time slices

```{r}
set.seed(32323, sample.kind = "Rounding")
tiempo <- 1:1000
folds <- createTimeSlices(y = tiempo, initialWindow = 20, horizon = 10)
names(folds)

folds$train[[1]]
folds$test[[1]]
```
# Training options

```{r, warning = FALSE}
library(caret)
library(kernlab)
data(spam)

inTrain <- createDataPartition(y = spam$type, p = 0.75, list = FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]

modelFit <- train(type ~., data = training, method = "glm")

args(train)
```

Among the metric options are:

- Continuous outcomes:
    + RMSE
    + R-Squared from regression models
- Categorical outcomes:
    + Accuraccy -> Fraction correct
    + Kappa -> A measure of concordance.
    
```{r}
args(trainControl)
```
Function `trainControl`

- Method
    + *boot*: bootstrapping
    + *boot632*: bootstrapping with adjustment
    + *cv*: Cross-validation
    + *repeatedcv*: repeated cross-validation
    + *LOOCV*: Leave one out cross-validation
- Number
    + For boot/cross validation
    + Number of subsamples to take
- repeats
    + Number of times to repeate subsampling
    + If big this can slow things down

 
```{r, warning = FALSE}
set.seed(1235)
modelFit2 <- train(type ~., data = training, method = "glm")
modelFit2
```

# Plotting predictors

Example: Predicting wages


```{r}
library(ISLR)
library(ggplot2)
library(caret)
data(Wage)
summary(Wage)
```
```{r}
inTrain <- createDataPartition(y = Wage$wage, p = 0.7, list = FALSE)
training <- Wage[inTrain,]
test <- Wage[-inTrain,]
dim(training)
dim(testing)
```

```{r}
featurePlot(x = training[,c("age", "education", "jobclass")],y = training$wage, plot = "pairs")
qplot(age,wage, data = training)
qplot(age, wage, data = training, colour = jobclass)
```

```{r}
qq <- qplot(age, wage, colour = education, data = training)
qq + geom_smooth(method = "lm", formula = y~x)
```
We can use the `cut` function to make factors. This function is included in `Hmisc` R package.

```{r}
library(Hmisc)
library(gridExtra)
cutWage <- cut2(training$wage, g = 3)
table(cutWage)
p1 <- qplot(cutWage, age, data = training, fill = cutWage, geom = c("boxplot"))
p1
p2 <- qplot(cutWage, age, data = training, fill = cutWage, geom = c("boxplot", "jitter"))
grid.arrange(p1,p2,ncol = 2)
```

We can create tables

```{r}
t1 <- table(cutWage, training$jobclass)
t1
### Use number 1 for row-wise proportion.
prop.table(t1,1)
```


```{r}
qplot(wage, colour = education, data = training, geom ="density")
```
Do not forget:

- Make your plots only **in the training set**: don't use the test set for exploration.
- Things you should be looking for:
    + Imbalance in outcomes/predictors
    + Outliers
    + Groups of points not explained by a predictor
    + Skewed variables
    
    
    
# Preprocessing

```{r}
library(caret)
library(kernlab)
data(spam)
inTrain <- createDataPartition(y = spam$type, p = 0.75, list = FALSE)

training <- spam[inTrain,]
testing <- spam[-inTrain,]
hist(training$capitalAve, main = "", xlab = "ave. capital run length")
mean(training$capitalAve)
sd(training$capitalAve)
```
This is a **very skewed** data. We can standardize. We can do it *manually* or use a function from the `caret` package, as it is shown in the following chunk.

```{r}
trainCapAve <- training$capitalAve
trainCapAveS <- (trainCapAve - mean(trainCapAve))/sd(trainCapAve)
c(mean(trainCapAveS), sd(trainCapAveS))

## Applied to the test set
testCapAve <- testing$capitalAve
testCapAveS <- (testCapAve - mean(trainCapAve))/sd(trainCapAve)
c(mean(testCapAveS), sd(testCapAveS))


## Using the caret function
preObj <- preProcess(training[,-58], method = c("center", "scale"))
trainCapAveS <- predict(preObj, training[,-58])$capitalAve
c(mean(trainCapAveS), sd(trainCapAveS))
testCapAveS <- predict(preObj, testing[,-58],)$capitalAve
c(mean(testCapAveS), sd(testCapAveS))
```


```{r, warning = FALSE}
set.seed(32343)
modelFit <- train(type ~., data = training, preProcess = c("center", "scale"), method = "glm")
modelFit
```


```{r}
preObj <- preProcess(training[,-58], method = c("BoxCox"))
trainCapAveS <- predict(preObj, training[,-58])$capitalAve
hist(trainCapAveS)
qqnorm(trainCapAveS)
```

## Standardizing - Imputing data

```{r}
set.seed(13343, sample.kind = "Rounding")

#Make some values NA
training$capAve <- training$capitalAve
selectNA <- rbinom(dim(training)[1], size = 1, prob = 0.05)==1
training$capAve[selectNA] <- NA

# Impute and standardize
preObj <- preProcess(training[,-58], method = "knnImpute")
capAve <- predict(preObj, training[,-58])$capAve

# Standardize true values
capAveTruth <- training$capitalAve
capAveTruth <- (capAveTruth - mean(capAveTruth))/sd(capAveTruth)

quantile(capAve - capAveTruth)
quantile((capAve - capAveTruth)[selectNA])
quantile((capAve - capAveTruth)[!selectNA])
```



Notes:

- Training and test must be processed in the same way.
- Test transformations will likely be imperfect, especially if the test/training sets collected at different times
- Careful when transforming *factor* variables!

# Covariate creation

Two levels of covariate creation: Level 1 is from raw data to covariates (they could be summary statistics of text, images or websites), and level 2 is transforming tidy covariates.

## Level 1: Raw data -> covariates

When doing this, take into account that:

- Examples of this include:
    + Text files: frequency of words, frequency of phrases, frequency of capital letters, etc.
    + Images: edges, corners, blobs, ridges.
    + Webpages: number and type of images, position of elements, colors, videos.
    + People: Height, weight, hair color, sex, country of origin
- Depends heavily on application.
- The balancing act is summarization vs. information loss
- The more knowledge of the system you have the better the job you will do
- When in doubt, err on the side of more features
- Can be automated, but use caution!


## Level 2: transform tidy covariates

- This is more necessary for some methods (such as regression and support vector machines) than for others (such as classification trees)
- Should be done **only** on the training set
- The best approach is through exploratory data analysis (plotting/tables)
- New covariates should be added to data frames


Let's take a look at one example:

```{r}
library(ISLR)
library(caret)
data(Wage)
inTrain <- createDataPartition(y = Wage$wage, p = 0.7, list = FALSE)
training <- Wage[inTrain,]
test <- Wage[-inTrain,]
```
A common practice is to convert qualitative variables into dummy variables

```{r}
table(training$jobclass)

dummies <- dummyVars(wage ~ jobclass, data = training)
head(predict(dummies, newdata = training))

```

Another common practice is to remove zero covariates

```{r}
nearzerocovariates <- nearZeroVar(training, saveMetrics = TRUE)
nearzerocovariates
```

Here, we can see that `region` has a near zero variability.


Spline basis can be constructed also:

```{r}
library(splines)
bsBasis <- bs(training$age, df =3)
bsBasis

lm1 <- lm(wage ~ bsBasis, data = training)
plot(training$age, training$wage, pch = 15, cex = 0.5)
points(training$age, predict(lm1, newdata = training), col = "red", pch = 19, cex =0.5)
```

Splines on the test set

```{r}
predict(bsBasis, age = testing$age)
```

## Notes and further reading

- Level 1 feature creation (raw data to covariates)
    + Science is key. Advice: (to) google "feature extraction for [data type]".
    + Err on overcreation of features
    + In some applications (images, voices) automated feature creation is possible/necessary
    + [Deep learning tutorial](https://cs.nyu.edu/~yann/talks/lecun-ranzato-icml2013.pdf)
- Level 2 feature creation (covariates to new covariates)
    + The function `preProcess` in `caret` R-package will handle some preprocessing
    + Create new covariates of you think they will improve the fit
    + Use EDA (Exploratory Data Analysis) on the training set for creating them
    + Be careful about overfitting!
- Preprocessing with caret: [Caret documentation](http://caret.r-forge.r-project.org/preprocess.html)
- In order to fit spline models, use the `gam` method in the `caret` package which allows smoothing of multiple variables


# Preprocessing with PCA

Let's take a look at **Principal Component Analysis, (PCA)**

### Correlated predictors

```{r}
library(caret)
library(kernlab)
data(spam)
inTrain <- createDataPartition(y = spam$type, p = 0.75, list = FALSE)

training <- spam[inTrain,]
testing <- spam[-inTrain,]

M <- abs(cor(training[,-58]))
diag(M) <- 0
which(M > 0.8, arr.ind = T)
```

```{r}
names(spam)[c(32,34,40)]
plot(spam[,32], spam[,34])
```
With this, we can see wich variables are highly correlated

### Basic PCA idea

- We might not need every predictor.
- A weighted combination of predictors might be better
- We should pick this combination to capture the *most information* possible
- Benefits:
    + Reduced number of predictors (Okham's razor?)
    + Reduced noise (due to averaging)

Using linear transformations (rotation matrix) we could rotate the plot 

```{r}
theta <- -pi/4

X <- cos(theta)*training$num857 - sin(theta)*training$num415
Y <- sin(theta)*training$num857 + cos(theta)*training$num415
plot(X,Y)
```
From the plot, $X$ captures more variation than $Y$, so that linear combination should be used

\[
    X = \left(\frac{1}{\sqrt{2}}\right)\text{num415} + \left(\frac{1}{\sqrt{2}}\right)\text{num857}
\]

### Related problems

You have multivariate variables $X_1, X_2, \dots, X_n$ so $X_1 = \left(X_{11}, \dots, X_{1m}\right)$

- Find a new set of multivariable variables that are **uncorrelated** and explain as much variance as possible
- If you put all the variables together in one matrix, find the best matrix created with fewer variables (**lower rank**) that explains the original data

The first goal is *statistical* and the second goal is *data compression*

### Related solutions - PCA/SVD

#### Singular Value Decomposition (SVD)

If $X$ is a matrix with each variable in a column and each observation in a row then the $SVD$ is a *matrix decomposition* 

\[
    X = UDV^T
\]

where the columns of $U$ are orthogonal (left singular vectors), the columns of $V$ are orthogonal (right singular vectors) and $D$ is a diagonal matrix (singular values)

#### Principal Components Analysis (PCA)

The principal components are equal to the right singular values **if** you first scale (substract the mean and divide by the standard deviation) the variables


Let's take a look at one example 

```{r}
library(kernlab)
data(spam)
smallSpam <- spam[,c(32,34)]
prComp <- prcomp(smallSpam)
plot(prComp$x[,1], prComp$x[,2])
prComp$rotation
```

Consider more variables: all the spam dataset.

```{r}
typeColor <- ((spam$type=="spam")*1+1)
prComp <- prcomp(log10(spam[,-58]+1))
plot(prComp$x[,1], prComp$x[,2], col = typeColor, xlab = "PC1", ylab = "PC2")
```

This can be done using `caret`

```{r}
preProc <- preProcess(log10(spam[,-58]+1), method = "pca", pcaComp = 2)
spamPC <- predict(preProc, log10(spam[,-58]+1))
plot(spamPC[,1], spamPC[,2], col = typeColor)
```

```{r, warning = FALSE}
preProc <- preProcess(log10(training[,-58]+1), method = "pca", pcaComp=2)
trainPC <- predict(preProc, log10(training[,-58]+1))
modelFit <- train(x = trainPC, y = training$type, method = "glm")

testPC <- predict(preProc, log10(testing[,-58]+1))
confusionMatrix(testing$type, predict(modelFit, testPC))
```
An alternative:

```{r, warning = FALSE}
modelFit2 <- train(type ~., method = "glm", preProcess = "pca", data = training)
confusionMatrix(testing$type, predict(modelFit2, testing))
```


Final thoughts on PC

- Most useful on linear-type models.
- Can make it harder to interpret predictors.
- Watch out for outliers!
    + Transform first (with logs/Box Cox)
    + Plot predictors to identify problems
- For more info see: [Elements of Statistical Learning](http://statweb.stanford.edu/~tibs/ElemStatLearn/)


# Predicting with regression

## Key ideas

- Fit a simple regression model
- Plug in new covariates and multiply by the coefficients
- Useful when the linear model is (nearly) correct.
- **Pros**
    + Easy to implement
    + Easy to interpret
- **Cons**
    + Often poor performance in non-linear settings
    

## Example: Old Faithful eruptions

```{r}
library(caret)
data(faithful)
set.seed(333)

inTrain <- createDataPartition(y = faithful$waiting, p = 0.5, list = FALSE)
trainFaith <- faithful[inTrain,]
testFaith <- faithful[-inTrain,]
head(trainFaith)
```

Check if a linear model is suitable

```{r}
plot(trainFaith$waiting, trainFaith$eruptions, pch = 19, col = "blue", xlab = "Waiting", ylab = "Duration")
```

```{r}
lm1 <- lm(eruptions ~ waiting, data = trainFaith)
summary(lm1)
```

```{r}
plot(trainFaith$waiting, trainFaith$eruptions, pch = 19, col = "blue", xlab = "Waiting", ylab = "Duration")
lines(trainFaith$waiting, lm1$fitted.values, lwd = 3)
```
Predict a new value: use the coefficients or use the `predict` function. For instance, assume $waiting = 80$

```{r}
coef(lm1)[1] + coef(lm1)[2]*80

newdata <- data.frame(waiting = 80)
predict(lm1,newdata)

newdata <- data.frame(waiting = c(20,60,80))
predict(lm1,newdata)
```

## Plot predictions - training and test

```{r}
par(mfrow = c(1,2))
plot(trainFaith$waiting, trainFaith$eruptions, pch = 19, col = "blue", xlab = "Waiting", ylab = "Duration")
lines(trainFaith$waiting, predict(lm1), lwd = 3)
plot(testFaith$waiting, testFaith$eruptions, pch = 19, col = "red", xlab = "Waiting", ylab = "Duration")
lines(testFaith$waiting, predict(lm1, newdata = testFaith), lwd = 3)
```
## Get the training and test set errors

```{r}
# Calculate the RMSE on training
sqrt(sum((lm1$fitted.values - trainFaith$eruptions)^2))
# Calculate the RMSE on test
sqrt(sum((predict(lm1,newdata = testFaith) - testFaith$eruptions)^2))

```

## Prediction intervals

```{r}
pred1 <- predict(lm1, newdata = testFaith, interval = "prediction")
ord <- order(testFaith$waiting)
plot(testFaith$waiting, testFaith$eruptions, pch = 19, col = "blue", xlab = "Waiting", ylab = "Duration")
matlines(testFaith$waiting[ord], pred1[ord,], type = "l", col = c("blue","red","red"), lty = c(1,1,1), lwd = 3)
```
## Same process with caret

```{r}
modelFit3 <- train(eruptions ~ waiting, data = trainFaith, method = "lm")
summary(modelFit3$finalModel)
```
## Notes and further reading

- Regression models with multiple covariates can be included
- Often useful in combination with other models
- [Elements of statistical learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
- [Modern applied statistics with S](http://www.amazon.com/Modern-Applied-Statistics-W-N-Venables/dp/0387954570)
- [Introduction to statistical learning](http://www-bcf.usc.edu/~gareth/ISL/)

# Predicting with regression multiple covariates

Let's take a look at one example. We use the `Wage` dataset and remove the variable `logwage` because we want to predict it using the rest of variables.

```{r}
library(ISLR)
library(ggplot2)
library(caret)
data(Wage)
Wage <- subset(Wage, select = -c(logwage))
summary(Wage)
```

Create the train and test sets

```{r}
inTrain <- createDataPartition(y = Wage$wage, p = 0.7, list = FALSE)
training <- Wage[inTrain,]
testing <- Wage[-inTrain,]
dim(training)
dim(testing)
```

```{r}
featurePlot(x = training[,c("age", "education", "jobclass")], y = training$wage, plot = "pairs")
qplot(age, wage, data = training, colour = jobclass)
qplot(age, wage, data = training, colour = education)
```
We can see that eduaction and job class might have effect on the variable `wage`. Therefore, we can fit a linear model. 

\[
    Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \left(\beta_3 X_3 + \beta_4 X_4 + \beta_5 X_5 + \beta_6 X_6\right)
\]

where:

- $Y$ is the `wage
- $X_1$ is the age
- $X_2, \dots, X_6$ are dummy variables:
    + $X_2 = 1$ if jobclass is information (therefore, $X_2= 0$ if joblcass is industrial)
    + $X_3 = 1$ if education level is
    + $X_4 = 1$ if education level is
    + $X_5 = 1$ if education level is
    + $X_6 = 1$ if education level is
    
```{r}
modFit <- train(wage ~ age + jobclass + education, method = "lm", data = training)
finMod <- modFit$finalModel
print(modFit)
print(finMod)
```
## Diagnostics

```{r}
plot(finMod, 1, pch=19, cex = 0.5, col = "#00000010")
```
We can see what happens when we colour by variables not included in the model:

```{r}
qplot(finMod$fitted.values, finMod$residuals, colour = race, data = training)
qplot(finMod$fitted.values, finMod$residuals, colour = health, data = training)
qplot(finMod$fitted.values, finMod$residuals, colour = maritl, data = training)
qplot(finMod$fitted.values, finMod$residuals, colour = health_ins, data = training)
```

## Plot residuals by index

```{r}
plot(finMod$residuals, pch = 19)
```

## Predicted versus truth in test set

```{r}
pred <- predict(modFit, testing)
qplot(wage, pred, colour = year, data = testing)
```

## Using all covariates

```{r}
modFitAll <- train(wage ~ ., data = training, method = "lm")
pred <- predict(modFitAll, testing)
qplot(wage, pred, data = testing)
```


